Transcriptional dynamics of influenza virus infection in single cells

### 2017-09-07 paper/paper.tex
Addressed comments and added rough Discussion section.

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
\jdbcomment{Make the matrix more in line with the data that we actually see. Make the counts of the different host genes reasonably similar between cells. We think most host genes have pretty similar expression across cells, but your example shows all having at least 2-fold difference and some much larger. So adjust the numbers to be more similar among cells. Also, in A and the matrix in C, just write out ``wildtype'', ``synonymous'', and ``not detected'' (or ``none``) -- there is plenty of space, so no need to use acronyms.}
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
\jdbcomment{Can you invert the y-axis on this plot so that it shows TCID50 / HA.}
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing the viral HA protein.
A549 cells were infected with all three viral stocks at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that the MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes generally becomes even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells \jdbcomment{What is the reference here?}.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\figdata{The sequence of the HA viral RNA carrying the GFP gene is in \texttt{HAflank-eGFP.fasta}.\label{figdata:HAflank}}
\end{figure}

Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
This observation is consistent with the finding by \citet{Dou:2017cp} and \citet{Huang:2008gy} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.

@@

\section{Discussion}
Still needs to be written.

@@

Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
The synonymous mutations were chosen to introduce nucleotides that are common in naturally occurring H1N1 viruses to avoid unwanted attenuation.
The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.

The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by \jdbcomment{Alistair, add this.}

@@

\subsection{qPCR}
For the qPCR in Figure~\ref{fig:viruspopulations} and Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection. 
On the day of infection, a single well was trypsinized and the cells were counted in order to determine the appropriate amount of virus to use to achieve the intended MOI.
Immediately before infection, D10 was replaced with influenza growth media.
For cells incubated with cyclohexamide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's instructions. 
cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
Transcript abundance was measured using SYBR\textsuperscript{TM} green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in \jdbcomment{just list the primers here, as there aren't that many of them}.
Biological triplicates were performed for all samples.
For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA \jdbcomment{what are these primers? If they are the same ones given above, indicate that. If there are from one of the references that are at the end of this sentence, then indicate that.} using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
qPCR was then performed as for mRNA measurements.
In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding reverse genetics plasmid. 
All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.

\subsection{Flow cytometry titering and analyses}
To determine viral titers in terms of HA-expressing units and for the flow cytometry in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
Cells were then washed, fixed in 1\% formaldehyde\jdbcomment{are you sure it is formaldehyde and not paramformaldehyde?}, and washed further before a final resuspension and analysis. 
To determine HA-expressing units, we then determined the fraction of cells that were HA positive and calculated the HA-expressing units from these fractions assuming a Poisson infection process.

For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating for single-cells in FlowJo\jdbcomment{What do you mean gating for single cells? The gates select populations of cells}, data were subsequently extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.

\subsection{Infections for single-cell mRNA sequencing}
All samples were processed using the 10x Chromium Single Cell 3' platform~\cite{zheng2017massively}, both in initial, tagged, cDNA generation and in final library preparation \jdbcomment{say the version of the reagents used -- v1 or v2? Also, in the preceding sentence, what do you mean by both the initial cDNA generation and in final library preparation? Can't you just say that you used 10X without specifying both?}.
All time points except for the 8-hour repeat, were performed on the same day.

>>>>

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing the viral HA protein.
A549 cells were infected with all three viral stocks at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that the MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes generally becomes even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells ~\citep{Heldt:2015iz}.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
{\bf(F)} Gaussian kernel density estimate on HA positive cells from {\bf(C)}
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D03.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\figdata{The sequence of the HA viral RNA carrying the GFP gene is in \texttt{HAflank-eGFP.fasta}.\label{figdata:HAflank}}
\end{figure} 
Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
This observation is consistent with the finding by \citet{Dou:2017cp} and \citet{Huang:2008gy} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.

@@

\section{Discussion}
We have generated total transcriptome measurements of individual influenza-infected cells -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
Many of our observations closely match those made by other methods, validating our approach and granting us confidence in exploring more nuanced features of our data.
Overall we find that influenza replication is defined broadly by both significant stochasticity and significant robustness, and that a view of viral infection as either strictly deterministic or incredibly chaotic fails to accurately describe the infectious process.
Moreover we found that stress, likely induced by oxidative damage, appears to define the cellular response to infection.

One of the most striking features of stochasticity with respect to influenza infection is the absence of genes corresponding to a given genomic segment, observed both by ourselves and others~\citep{Brooke:2013kb,Dou:2017cp}
While there is significant speculation, and some indirect evidence, regarding the step of viral replication at which these genes are lost, there currently exists no strong consensus as to its identity -- indeed there may be no single event wholly responsible the stochastic loss of viral gene expression ~\citep{Brooke:2017gm, Heldt:2015iz}
Regardless it does not seem unreasonable to surmise that this is a consequence of possessing a segmented genome. If so, it can be reasonably concluded that at some phase of viral replication, whether packaging, delivery to the nucleus, production of replicative intermediates, or some as-yet unsurmised step, the components of the viral genome must act in at least a semi-independent fashion. 
A consequence of this stochasticity is a subpopulation of infected cells that express significantly lower levels of viral transcripts. 
Consistent with prior population-level studies of transcription, we find that absence of any of the four viral gene products necessary to form the viral polymerase leads to a dramatic drop in the levels of mRNA transcripts detected~\citep{Vreede:2004ip,eisfeld2015centre}. 
Our results are in slight contrast to a prior flow cytometric analysis of protein expression that described a relatively minor drop in expression associated with absence of viral ribonucleoprotein~\citep{Brooke:2013kb}.
This discrepancy could be explained by two possibilities. 
The first is that mRNA levels are not wholly predictive of viral protein abundance. 
The second is that flow cytometry depends on a user-defined dynamic range of sensitivity as well as positive and negative gates, and that the potential 2-3 log difference in population means was either not captured in the sensitivity of the instrument or that gates on positive populations excluded bone-fide positives. 
Indeed, gating on HA staining using a strict cutoff based on an uninfected population alone produces a curve that is consistent with a bimodal distribution of HA expression (Figure ~\ref{fig:coexpression} Figure Supplement ~\ref{figsupp:flowcyto}F).

However, despite extensive heterogeneity, there seems to be robustness to the infectious lifecycle.
For example, influenza, regardless of the viral burden, exhibits relatively constrained ratios between the viral transcripts.
This suggests that the processes that are driving heterogeneity of influenza mRNA abundance act equivalently throughout the viral genome.
Potential mechanisms include components such as binding and entry kinetics that act on a virion and thus on all segments packaged therein, limitations of transcription rate that are dependent upon the same pool of cellular resources, or potentially even feedback-dependent mechanisms controlling viral transcription.

Perhaps even more important to influenza?s success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events. 
While non-infectious events might be considered metabolic ?waste?, they would, theoretically, not impact the growth of successful viral infections. 
However, recognition of any virus, infectious or not, and production of interferon, would impact the entire viral population. 
While we cannot extrapolate much from our single interferon-positive cell, it can be stated that interferon induction by influenza is rare in our dataset. 
This is consistent with prior experiments using reporter cell lines,  which places the induction at ~1\% of cells with detectable influenza protein expression -- albeit this value is likely sensitive to differences in viral propagation and strain utilized ~\citep{Killip:2017ef, Chen:2010cr}. 
If the major influenza interferon antagonist, NS1, is absent in ~10-20\% of infections, and lowly expressed in those infections engaging in only primary transcription, we must posit that NS1 alone is insufficient to explain influenza evasion of innate immunity. 
Interestingly, if other segmented viruses also exhibit a pattern of stochastic segment loss, they too would need to overcome this hurdle and escape innate immunity through mechanisms relying on more than a gene product generated by a single segment.

In addition to viral mRNA kinetics, our analysis also allows us to observe those features of the cellular transcriptome that covary with influenza transcription.
Critically, our analysis does not require synchronization or infection with a high MOI, and thus avoids potential artifacts produced by either approach. 
We can therefore see extreme changes in a small subset of cells within a larger population.
Moreover, we can also observe how transcriptional changes partition between cells, and exclude dramatic changes such as interferon induction that, while important, can mask other interesting biologically-relevant responses. 
With this in mind, the predominant response associated with increased influenza mRNA transcript abundance appears to be the oxidative stress pathway. 
Generally characterized as antiviral, it is possible that cellular stress pathways represent an additional arm of innate immunity -- detecting a cellular state consistent with a loss of homeostasis as resources are diverted to viral propagation. 
Intriguingly, one element of the response, cyclophilin A (\emph{CypA}), is both thought to suppress influenza replication through degradation of matrix protein as well as act as a proinflammatory signal ~\citep{Liu:2009er, Jin:2004ii}.
It will be interesting to follow up on these observations and determine the broader mechanisms through which cellular stress pathways are involved in pro- or anti-viral responses.

The success garnered here underscores that single-cell approaches seem distinctly suited to study of viral pathogens, as not only are many host responses governed by bimodality and stochasticity ~\citep{Shalek:2014ey,Shalek:2013ej}, but also viral populations themselves exhibit a staggering array of diversity.
Our results with influenza have provided significant support for earlier analysis of viral stochasticity, further defined the sources and impacts of that stochasticity, and generated potentially fertile avenues of research. 
While it is early days yet, it is our hope that our efforts will contribute to a deeper understanding of viral populations, treating population-level measurements as emergent phenomena defined by underlying individual stochasticity rather than analyzing viral replication in a strictly deterministic framework.

@@

Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
The synonymous mutations were chosen to introduce nucleotides that are common in naturally occurring H1N1 viruses to avoid unwanted attenuation.
The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours, when obvious cytopathic effect was noted.

@@

\subsection{qPCR}
For the qPCR in Figure~\ref{fig:viruspopulations} and Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection. 
On the day of infection, a single well was trypsinized and the cells were counted in order to determine the appropriate amount of virus to use to achieve the intended MOI.
Immediately before infection, D10 was replaced with influenza growth media.
For cells incubated with cyclohexamide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's instructions. 
cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
Transcript abundance was measured using SYBR\textsuperscript{TM} green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the following primers: \emph{HA}: 5'-GGCCCAACCACACATTCAAC-3', 5'-GCTCATCACTGCTAGACGGG-3', \emph{IFNB1}: 5'-AAACTCATGAGCAGTCTGCA-3', 5'-AGGAGATCTTCAGTTTCGGAGG-3', \emph{L32}: 5'-AGCTCCCAAAAATAGACGCAC-3', 5'-TTCATAGCAGTAGGCACAAAGG-3'.   
Biological triplicates were performed for all samples.
For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA adapted from \citep{Hoffmann:2001vj}  using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl}.
qPCR was then performed as for mRNA measurements.
In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding reverse genetics plasmid. 
All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.

\subsection{Flow cytometry titering and analyses}
To determine viral titers in terms of HA-expressing units and for the flow cytometry in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
To determine HA-expressing units, we then determined the fraction of cells that were HA positive and calculated the HA-expressing units from these fractions assuming a Poisson infection process.

For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating to exclude multiplets in FlowJo, data were subsequently extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.

\subsection{Infections for single-cell mRNA sequencing}
Single cell sequencing libraries were generated using the 10x Chromium Single Cell 3' platform, V1~\cite{zheng2017massively}.
All time points except for the 8-hour repeat, were performed on the same day.
